### 2020-05-22 main.Rmd
First draft of methods and results

@@

We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs. Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in table \@ref(tab:overall_model). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:model_corrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 - 0.9; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 - 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.3 - 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

@@

We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input. We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Table \@ref(tab:ensembleresults). With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions. However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures. The

## Comparison With Other Classification Software
Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models to evaluate the performance of our models against widely used effector identification software

^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document]

@@

@@

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Effectors found in three major pathogen groups'
)
```

@@

```{r, ensembleresults}
apa_table(
  readr::read_tsv("tables/ensemble_model_performance.tab"),
  digits = 3,
  align = 'c',
  caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
)
```

```{r, overallmodel}
apa_table(
  readr::read_tsv("tables/final_model_performance.tab"),
  digits = 3,
  align = 'c',
  caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
)
```

>>>>

F1-score is the harmonic average of the precision and recall:
\begin{equation}
  F_{1} = \frac{2TP}{2TP + FN + FP}
\end{equation}

## Activation Map Analysis

To visualise the regions of the sequences that the models were using to discriminate between effector and non-effector we used a GRAD-CAM approach ^[Ruth, there is absolutely no description of this in your report, please add a description of what you did, with which inputs and settings and versions etc].

@@

We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs. Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4 % to 97.4 %, for the four fungal models we observed accuracy in the range 60.5 % to 84.2 % and for the four oomycete models we saw accuracy from 64.7 % to 82.3 %, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

@@

Classification correlations between the different model architectures were high and positive in the bacterial sequence trained model's calls, in the range 0.8 - 0.88, see Figure \@ref(fig:modelcorrs). CNN-GRU and LSTM-Embedding showed identical prediction sets. We saw similar levels of correlation  in the CNN-LSTM, GRU-Embedding and CNN-GRU fungal sequence trained model, in the range 0.79 - 0.9; though there was a significantly lower range of correlations with the LSTM-Embedding which were in the range 0.36 - 0.51. The models trained on oomycete sequences all showed this lower range of correlations, in the range 0.3 - 0.65. The higher correlation across bacterial trained models is likely from a mixture of the larger training set size and a greater uniformity of the sequences themselves. For the fungal sequence trained models we can see that the LSTM-Embedding model does not perform as well as the others. The oomycete sequence trained models all show a lower range correlation reflecting the likely less uniform and smaller training set. It is clear that, particularly for the fungal and oomycete models, each architecture is capturing separate aspects of the sequences and classifying on those with slightly varying levels of success.

@@

We examined the usefulness of combining the predictions of the different model architectures using an ensemble function that takes the vectors of classifications of each model architecture as input. We performed the classification of the hold-out test data set using the ensembled models and the results can be seen in Figure \@ref(fig:ensembleresults). With the models trained in bacterial sequences we saw an increase in classification accuracy over the best model, up to 0.99 for both ensemble functions. However, with the fungal and oomycete models we saw decreases relative to the best single model in both cases due to the higher accuracy of the CNN-LSTM model being diluted by the combined inacuracy of the other model architectures. Examining the overlaps in classifications between the CNN-LSTM and CNN-GRU/LSTM-Embedding respectively showed that the two lesser performing models were not simply predicting subsets of the CNN-LSTM model, in both cases the lesser models were able to identify three effectors correctly that were missed by the generally stronger models. This indicates that the weaker models may be classifying on some patterns missed by the CNN-LSTM model.

## Comparison With Other Classification Software

Given the accuracies of the above we selected the ensemble bacterial model and the CNN-LSTM fungal and oomycete models ^[Ruth, can you confirm that this is accurate for the rest of the analysis - are these the combinations of models you are using: You will need to discuss this with me before we make any changes to this document] to evaluate the performance of our models against widely used effector identification software. We compared against predictions from the bacterial effector prediction programs deepT3 and effectiveT3, the fungal effector prediction programs EffectorP 1 and 2 and the oomycete effector predictor effectR. Each comparison was carried out using the respective hold-out test sequence set for each taxon. For all taxon we saw greater Accuracy and F1 scores from our models than the established software, as shown in Figure \@ref(fig:vsothersmetrics). This was particularly marked in the F1 score, which incorporates a measure of the incorrect calls. Absoloute improvements were up to ~ 18 % in bacterial sequences, ~15 % in fungal sequences and ~ 28% in the oomycete sequences. The confusion matrices in Figure \@ref(fig:vsotherscfmat) show that accuracy and F1 score was compromised in all the established tools by the tendency of them all to misclassify true effectors as not effectors.  All the established software classifiers we tested show lower sensitivity than the models we have developed here.

@@

An advantage of CNNs relative to other deep neural networks is their relative interpretability. A CNN can be analysed to activation maps which  highlight the regions of the input to which the CNN is most strongly using to classify. To examine the responses of the model, we ran the effector (not the non-effector) ^[Ruth, can you confirm that this is what you did, we want the effector activations only, not the whole set, or at least the effector activations in one panel and the non-effectors in another]. data sets back through the CNN-LSTM models for each taxon and extracted the network activations using the GRAD-CAM method , the profiles were smoothed using FFT and examined \@ref(fig:activations). All the models showed a peak of activation at the N-terminus of the sequences, coincident with expected positions of secretion signals. The fungal sequences created a single broad activation region with a width of 50 to 100 amino acids while the bacterial and oomycete sequences create a some smaller grouped peaks in a broader region which were each around 20 amino acids. Compositional analysis of the sequence under the peaks showed no apparent primary sequence conservation or motifs even within a taxon and data set. The sequences in the activation regions in all the taxa and models we examined did seem to be preferentially made up of hydrophobic and polar amino acids, which is apparently contradictory ^[Ruth, we need to discuss this bit again, we may need to re-run on individual sets, ie training sets with just effectors in...]

@@

To make the models useful for developers in their own analytic pipelines we have provided an R package that provides a useful interface to the code. The package is available from GitHub [https://ruthkr.github.io/deepredeff](https://ruthkr.github.io/deepredeff)

^[Ruth, will we do a shiny tool?]

```{r, effsummary}
apa_table(
  readr::read_tsv("tables/effectors_by_path_group.tab"),
  digits = 0,
  align = 'c',
  caption = 'Count of effectors listed in publications curated by PHI-Base used in this study in three major plant pathogen groups'
)
```

```{r, overallmodel, fig.cap="Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/final_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of models in each taxon on the unseen, hold-out test data fraction of sequences.'
# )
readr::read_tsv("tables/final_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Model, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = Accuracy)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")

```{r, ensembleresults, fig.cap="Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences."}
# apa_table(
#   readr::read_tsv("tables/ensemble_model_performance.tab"),
#   digits = 3,
#   align = 'c',
#   caption = 'Final performance of Ensemble models in each taxon on the unseen, hold-out test data fraction of sequences.'
readr::read_tsv("tables/ensemble_model_performance_tidy.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Combination, Accuracy) +
  ggplot2::geom_point(ggplot2::aes(colour = EnsembleFunction)) +
  ggplot2::facet_wrap(~ Taxon, scales = "free_x")

```{r, vsothersmetrics, fig.cap="Accuracy and F1 score of the classifications made by our models compared with those made by widely used software"}
readr::read_tsv("tables/cfothertools.tab") %>% 
  ggplot2::ggplot() +
  ggplot2::aes(Classifier, Score) +
  ggplot2::geom_point(ggplot2::aes(colour = Metric)) +
  ggplot2::facet_wrap(~ Group, scales = "free_x")
```

```{r, vsotherscfmat, fig.cap="Confusion matrices of classifications made by our models compared with those made by widely used software. A: Tools evaluated on bacterial sequences; B: Tools evaluated on fungal sequences; C: Tools evaluated on oomycete sequences."}
knitr::include_graphics("figures/cf_other_tools.png")
```

```{r, activations, fig.cap="GRAD-CAM activation profiles of CNN-LSTM models trained on when run with effector sequence data and sequence logo plots of sequence in the activation region within the dotted lines. A: Models trained on bacterial sequences; B: Models trained on fungal sequences; C: Models trained on oomycete sequences."}
knitr::include_graphics("figures/convolution_maps.png")
```

# Supplementary Tables

@@
